The locally Gaussian partial correlation

### 2020-09-16 lgpc_JBES.Rmd
New set of comments from Dag

An integral part of our article is the online supplement. It deals with a number of issues complementing the treatment in the main paper. Among them are more examples, extensions of the main concept to more general situations, precise formulations of some theoretical results with proofs, as well as discussions of local asymptotic power and the Pitman criterion for testing and validity of the use of the bootstrap.

@@

In the following, we will not always distinguish between $Z_j$ and $\widehat Z_{j}$. In fact, we conclude our asymptotic analysis in Section \@ref(chap:asymptotic) by showing that under certain technical conditions, the error made by estimating $\R(\z)$ using the empirically transformed variables $Z_{j,n}$, instead of $Z_j$, is smaller in the limit than the estimation error made when estimating the local correlations themselves.

@@

To see that conditional independence between $X_1$ and $X_2$ given $\X_3$ is equivalent to $\alpha(\z) \equiv 0$, or equivalently $\alpha(\x) \equiv 0$, we need to follow a few simple steps that would also work in the global Gaussian case. Working on the standard normal $z$-scale and assuming that $(z_1, z_2, \z_3)$ has density function $f_{\Z}(\z)$ as given by \@ref(eq:lgdeapprox), it follows from @otneim2017conditional that we can calculate conditional distributions in the same way as in the global Gaussian case. For example, the conditional density of $Z_1|\Z_3$ at the point $z_1$ is given by

@@

In Figure \@ref(fig:gaussian-example-plots) we see the estimated LGPC between $X_1$ and $X_2$ given $X_3=0$ for three samples, having sample size 100, 500 and 2000 respectively, mapped out on a grid. In this and all other examples we select bandwidths based on a simple plug-in formula that follows naturally from classical asymptotic arguments, $b = cn^{-1/9}$ for the full trivariate fit, and $b = cn^{-1/6}$ for the bivariate simplification, where the constant $c$ controls the amount of smoothing and must be chosen appropriately based on the task at hand. @otneim2016multivariate argues that $c=1.75$ is a good choice for multivariate density estimation, and we will in the next section see that we obtain good power in our test for conditional independence if $c$ is somewhat smaller than that. For the visual display of conditional dependence maps, however, we tend to prefer a fair amount of smoothing, and in Figure \@ref(fig:gaussian-example-plots) we have used $c=4$. @otneim2017locally also suggest a algorithm for bandwidth selection in the density estimation context, maximizing the leave-one-out cross validated log-likelihood of the sample;
$$\bb_{CV} = \textrm{arg max}_{\bb} \sum_{t = 1}^n \widehat{f}_{\bb}^{(-t)}(\Z_t),$$
where $\widehat f_{\bb}^{(-t)}(\Z_t)$ is the estimated local Gaussian density estimate calculated without, and evaluated in, the sample point $\Z_t$, using the bandwidth matrix $\bb$. A theory of efficient and accurate bandwidth selection for local likelihood estimation remains a topic for future research.

@@

We restrict ourselves to the case where $X_1$ and $X_2$ are scalars, since this is the situation considered in the papers we compare with, but we reiterate (end of Section \@ref(chap:definition)) that there are no conceptual difficulties in extending to the vector case as is seen in the online supplement. The theory underlying LGPC testing is essentially based on @lacal2018estimating, and is outlined in Sections F--H of the supplement.

@@

The first set of simulation results are collected from @su2008nonparametric: In Table \ref{tab:n100} LIN is a standard linear Granger causality test, SCM and SKS are tests by @linton1997conditional that use statistics of the two versions of the nonparametric test developed by @delgado2001significance. HEL refers to the Hellinger distance test that @su2008nonparametric present. Next, we move to @su2014testing, who provide simulations for their test for conditional independence based on the empirical likelihood (SEL), as well as the test by @su2007consistent that is based on properties of the conditional characteristic function (CHF). @cheng2012conditional provide simulations for their tests based on the maximal conditional correlation (MCC). In the online supplement to this article we provide a richer version of Table \@ref(tab:n100) below, namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.

@@

For 3 scalar variables $X_1, X_2, X_3$ a full trivariate analysis depending on all 3 coordinates can be undertaken as in Section \@ref(chap:trivariate-full). If any of the components $X_1, X_2$ or $X_3$ are vectors, a pairwise approximation and simplification has been described in Section \@ref(chap:bivariate). This can be likened to additive approximation in nonparametric regression.

>>>>

An integral part of our article is the online supplement. It deals with a number of issues complementing the treatment in the main paper. Among them are more examples, extensions of the main concept to more general situations, precise formulations of some theoretical results with proofs, as well as discussions of local asymptotic power and the Pitman criterion for testing, and validity of the use of the bootstrap.

@@

In the following, we will not always distinguish between $Z_j$ and $\widehat Z_{j}$. In fact, we conclude our asymptotic analysis in Section \@ref(chap:asymptotic) by showing that under certain technical conditions, the error made by estimating $\R(\z)$ using the empirically transformed variables $\widehat Z_{j}$, instead of $Z_j$, is smaller in the limit than the estimation error made when estimating the local correlations themselves.

@@

To see that conditional independence between $X_1$ and $X_2$ given $\X_3$ is equivalent to $\alpha(\z) \equiv 0$, or equivalently $\alpha(\x) \equiv 0$, we need to follow a few simple steps that would also work in the global Gaussian case. Working on the standard normal $z$-scale and assuming that $(Z_1, Z_2, \Z_3)$ has density function $f_{\Z}(\z)$ as in \@ref(eq:lgdeapprox), it follows from @otneim2017conditional that we can calculate conditional distributions in the same way as in the global Gaussian case. For example, the conditional density of $Z_1|\Z_3$ at the point $(z_1, \z_3)$ is given by

@@

In Figure \@ref(fig:gaussian-example-plots) we see the estimated LGPC between $X_1$ and $X_2$ given $X_3=0$ for three samples, having sample size 100, 500 and 2000 respectively, mapped out on a grid. In this and all other examples we select bandwidths based on a simple plug-in formula that follows naturally from classical asymptotic arguments, $b = cn^{-1/9}$ for the full trivariate fit, and $b = cn^{-1/6}$ for the bivariate simplification, where the constant $c$ controls the amount of smoothing and must be chosen appropriately based on the task at hand. @otneim2016multivariate argues that $c=1.75$ is a good choice for multivariate density estimation, and we will in the next section see that we obtain good power in our test for conditional independence if $c$ is somewhat smaller than that. For the visual display of conditional dependence maps, however, we tend to prefer a fair amount of smoothing, and in Figure \@ref(fig:gaussian-example-plots) we have used $c=4$. @otneim2017locally also suggest an algorithm for bandwidth selection in the density estimation context, maximizing the leave-one-out cross validated log-likelihood of the sample;
$$\bb_{CV} = \textrm{arg max}_{\bb} \sum_{t = 1}^n \widehat{f}_{\bb}^{(-t)}(\Z_t),$$
where $\widehat f_{\bb}^{(-t)}(\Z_t)$ is the estimated local Gaussian density estimate calculated without, and evaluated in, the sample point $\Z_t$, using the bandwidth matrix $\bb$. A theory of efficient and accurate bandwidth selection for local likelihood estimation remains a topic for future research.

@@

We restrict ourselves to the case where $X_1$ and $X_2$ are scalars, since this is the situation considered in the papers we compare with, but we reiterate (end of Section \@ref(chap:definition)) that there are no conceptual difficulties in extending to the vector case as is seen in the online supplement. The theory underlying LGPC testing is essentially based on @lacal2018estimating, and is outlined for the full vector case in Sections F--H of the supplement.

@@

The first set of simulation results are collected from @su2008nonparametric: In Table \ref{tab:n100} LIN is a standard linear Granger causality test, SCM and SKS are tests by @linton1997conditional that use statistics of the two versions of the nonparametric test developed by @delgado2001significance. HEL refers to the Hellinger distance test that @su2008nonparametric present. Next, we move to @su2014testing, who provide simulations for their test for conditional independence based on the empirical likelihood (SEL), as well as the test by @su2007consistent that is based on properties of the conditional characteristic function (CHF). @cheng2012conditional provide simulations for their tests based on the maximal conditional correlation (MCC). In the online supplement to this article we provide a richer version of Table \@ref(tab:n100), namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.

@@

For 3 scalar variables $X_1, X_2, X_3$ a full trivariate analysis depending on all 3 coordinates can be undertaken as in Section \@ref(chap:trivariate-full). If any of the components $X_1, X_2$ or $X_3$ are vectors, a pairwise approximation and simplification has been described in Section \@ref(chap:bivariate) and in the supplement. This can be likened to additive approximation in nonparametric regression.

### 2020-09-16 lgpc_JBES.tex
New set of comments from Dag

In the following, we will not always distinguish between \(Z_j\) and \(\widehat Z_{j}\). In fact, we conclude our asymptotic analysis in Section \ref{chap:asymptotic} by showing that under certain technical conditions, the error made by estimating \(\R(\z)\) using the empirically transformed variables \(Z_{j,n}\), instead of \(Z_j\), is smaller in the limit than the estimation error made when estimating the local correlations themselves.

@@

To see that conditional independence between \(X_1\) and \(X_2\) given \(\X_3\) is equivalent to \(\alpha(\z) \equiv 0\), or equivalently \(\alpha(\x) \equiv 0\), we need to follow a few simple steps that would also work in the global Gaussian case. Working on the standard normal \(z\)-scale and assuming that \((z_1, z_2, \z_3)\) has density function \(f_{\Z}(\z)\) as given by \eqref{eq:lgdeapprox}, it follows from Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) that we can calculate conditional distributions in the same way as in the global Gaussian case. For example, the conditional density of \(Z_1|\Z_3\) at the point \(z_1\) is given by

@@

Furthermore, Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) show that an analytic expression of the leading term in the covariance matrix \(\Jb^{-1}\Mb(\Jb^{-1})^T\) is not practically available. In applications we may circumvent this problem by using the bootstrap. The validity of the bootstrap can be proved using a direct adaptations of the techniques of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}). It is spelled out in some detail for the pairwise case in Theorems E.1, E.2 and corollaries E.3 and E.4 in Section E of the online supplement. The same arguments can be used in the present 3-parameter case (in fact, Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) treat the more general five-parameter case).

@@

In Figure \ref{fig:gaussian-example-plots} we see the estimated LGPC between \(X_1\) and \(X_2\) given \(X_3=0\) for three samples, having sample size 100, 500 and 2000 respectively, mapped out on a grid. In this and all other examples we select bandwidths based on a simple plug-in formula that follows naturally from classical asymptotic arguments, \(b = cn^{-1/9}\) for the full trivariate fit, and \(b = cn^{-1/6}\) for the bivariate simplification, where the constant \(c\) controls the amount of smoothing and must be chosen appropriately based on the task at hand. Otneim (\protect\hyperlink{ref-otneim2016multivariate}{2016}) argues that \(c=1.75\) is a good choice for multivariate density estimation, and we will in the next section see that we obtain good power in our test for conditional independence if \(c\) is somewhat smaller than that. For the visual display of conditional dependence maps, however, we tend to prefer a fair amount of smoothing, and in Figure \ref{fig:gaussian-example-plots} we have used \(c=4\). Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) also suggest a algorithm for bandwidth selection in the density estimation context, maximizing the leave-one-out cross validated log-likelihood of the sample;
\[\bb_{CV} = \textrm{arg max}_{\bb} \sum_{t = 1}^n \widehat{f}_{\bb}^{(-t)}(\Z_t),\]
where \(\widehat f_{\bb}^{(-t)}(\Z_t)\) is the estimated local Gaussian density estimate calculated without, and evaluated in, the sample point \(\Z_t\), using the bandwidth matrix \(\bb\). A theory of efficient and accurate bandwidth selection for local likelihood estimation remains a topic for future research.

@@

We restrict ourselves to the case where \(X_1\) and \(X_2\) are scalars, since this is the situation considered in the papers we compare with, but we reiterate (end of Section \ref{chap:definition}) that there are no conceptual difficulties in extending to the vector case as is seen in the online supplement. The theory underlying LGPC testing is essentially based on Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), and is outlined in Sections F--H of the supplement.

@@

Several authors, for example Teräsvirta, Tjøstheim, and Granger (\protect\hyperlink{ref-terasvirta2010modelling}{2010}), have noted, however, that asymptotic analysis of nonparametric (or semiparametric) test statistics on the form \eqref{eq:teststatistic} tends to be too crude for finite sample applications. We have therefore approximated the distribution of \(T_{n,b}\) using the bootstrap. The validity of the bootstrap in such contexts has been examined by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) and we adapt their arguments to our situation in the online supplement, Section H.

@@

The first set of simulation results are collected from Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}): In Table \ref{tab:n100} LIN is a standard linear Granger causality test, SCM and SKS are tests by Linton and Gozalo (\protect\hyperlink{ref-linton1997conditional}{1997}) that use statistics of the two versions of the nonparametric test developed by Delgado and Manteiga (\protect\hyperlink{ref-delgado2001significance}{2001}). HEL refers to the Hellinger distance test that Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}) present. Next, we move to Su and White (\protect\hyperlink{ref-su2014testing}{2014}), who provide simulations for their test for conditional independence based on the empirical likelihood (SEL), as well as the test by Su and White (\protect\hyperlink{ref-su2007consistent}{2007}) that is based on properties of the conditional characteristic function (CHF). Cheng and Huang (\protect\hyperlink{ref-cheng2012conditional}{2012}) provide simulations for their tests based on the maximal conditional correlation (MCC). In the online supplement to this article we provide a richer version of Table \ref{tab:n100} below, namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.

@@

For 3 scalar variables \(X_1, X_2, X_3\) a full trivariate analysis depending on all 3 coordinates can be undertaken as in Section \ref{chap:trivariate-full}. If any of the components \(X_1, X_2\) or \(X_3\) are vectors, a pairwise approximation and simplification has been described in Section \ref{chap:bivariate}. This can be likened to additive approximation in nonparametric regression.

@@

\leavevmode\hypertarget{ref-lacal2018estimating}{}%
---------. 2018. ``Estimating and Testing Nonlinear Local Dependence Between Two Time Series.'' \emph{Journal of Business \& Economic Statistics}, 1--13.

>>>>

In the following, we will not always distinguish between \(Z_j\) and \(\widehat Z_{j}\). In fact, we conclude our asymptotic analysis in Section \ref{chap:asymptotic} by showing that under certain technical conditions, the error made by estimating \(\R(\z)\) using the empirically transformed variables \(\widehat Z_{j}\), instead of \(Z_j\), is smaller in the limit than the estimation error made when estimating the local correlations themselves.

@@

To see that conditional independence between \(X_1\) and \(X_2\) given \(\X_3\) is equivalent to \(\alpha(\z) \equiv 0\), or equivalently \(\alpha(\x) \equiv 0\), we need to follow a few simple steps that would also work in the global Gaussian case. Working on the standard normal \(z\)-scale and assuming that \((Z_1, Z_2, \Z_3)\) has density function \(f_{\Z}(\z)\) as in \eqref{eq:lgdeapprox}, it follows from Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) that we can calculate conditional distributions in the same way as in the global Gaussian case. For example, the conditional density of \(Z_1|\Z_3\) at the point \((z_1, \z_3)\) is given by

@@

Furthermore, Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) show that an analytic expression of the leading term in the covariance matrix \(\Jb^{-1}\Mb(\Jb^{-1})^T\) is not practically available. In applications we may circumvent this problem by using the bootstrap. The validity of the bootstrap can be proved using a direct adaptations of the techniques of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}). It is spelled out in some detail for the pairwise case in Theorems E.1, E.2 and corollaries E.3 and E.4 in Section E of the online supplement. The same arguments can be used in the present 3-parameter case (in fact, Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) treat the more general five-parameter case).

@@

In Figure \ref{fig:gaussian-example-plots} we see the estimated LGPC between \(X_1\) and \(X_2\) given \(X_3=0\) for three samples, having sample size 100, 500 and 2000 respectively, mapped out on a grid. In this and all other examples we select bandwidths based on a simple plug-in formula that follows naturally from classical asymptotic arguments, \(b = cn^{-1/9}\) for the full trivariate fit, and \(b = cn^{-1/6}\) for the bivariate simplification, where the constant \(c\) controls the amount of smoothing and must be chosen appropriately based on the task at hand. Otneim (\protect\hyperlink{ref-otneim2016multivariate}{2016}) argues that \(c=1.75\) is a good choice for multivariate density estimation, and we will in the next section see that we obtain good power in our test for conditional independence if \(c\) is somewhat smaller than that. For the visual display of conditional dependence maps, however, we tend to prefer a fair amount of smoothing, and in Figure \ref{fig:gaussian-example-plots} we have used \(c=4\). Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) also suggest an algorithm for bandwidth selection in the density estimation context, maximizing the leave-one-out cross validated log-likelihood of the sample;
\[\bb_{CV} = \textrm{arg max}_{\bb} \sum_{t = 1}^n \widehat{f}_{\bb}^{(-t)}(\Z_t),\]
where \(\widehat f_{\bb}^{(-t)}(\Z_t)\) is the estimated local Gaussian density estimate calculated without, and evaluated in, the sample point \(\Z_t\), using the bandwidth matrix \(\bb\). A theory of efficient and accurate bandwidth selection for local likelihood estimation remains a topic for future research.

@@

We restrict ourselves to the case where \(X_1\) and \(X_2\) are scalars, since this is the situation considered in the papers we compare with, but we reiterate (end of Section \ref{chap:definition}) that there are no conceptual difficulties in extending to the vector case as is seen in the online supplement. The theory underlying LGPC testing is essentially based on Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}), and is outlined for the full vector case in Sections F--H of the supplement.

@@

Several authors, for example Teräsvirta, Tjøstheim, and Granger (\protect\hyperlink{ref-terasvirta2010modelling}{2010}), have noted, however, that asymptotic analysis of nonparametric (or semiparametric) test statistics on the form \eqref{eq:teststatistic} tends to be too crude for finite sample applications. We have therefore approximated the distribution of \(T_{n,b}\) using the bootstrap. The validity of the bootstrap in such contexts has been examined by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) and we adapt their arguments to our situation in the online supplement, Section H.

@@

The first set of simulation results are collected from Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}): In Table \ref{tab:n100} LIN is a standard linear Granger causality test, SCM and SKS are tests by Linton and Gozalo (\protect\hyperlink{ref-linton1997conditional}{1997}) that use statistics of the two versions of the nonparametric test developed by Delgado and Manteiga (\protect\hyperlink{ref-delgado2001significance}{2001}). HEL refers to the Hellinger distance test that Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}) present. Next, we move to Su and White (\protect\hyperlink{ref-su2014testing}{2014}), who provide simulations for their test for conditional independence based on the empirical likelihood (SEL), as well as the test by Su and White (\protect\hyperlink{ref-su2007consistent}{2007}) that is based on properties of the conditional characteristic function (CHF). Cheng and Huang (\protect\hyperlink{ref-cheng2012conditional}{2012}) provide simulations for their tests based on the maximal conditional correlation (MCC). In the online supplement to this article we provide a richer version of Table \ref{tab:n100}, namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.

@@

For 3 scalar variables \(X_1, X_2, X_3\) a full trivariate analysis depending on all 3 coordinates can be undertaken as in Section \ref{chap:trivariate-full}. If any of the components \(X_1, X_2\) or \(X_3\) are vectors, a pairwise approximation and simplification has been described in Section \ref{chap:bivariate} and in the supplement. This can be likened to additive approximation in nonparametric regression.

@@

\leavevmode\hypertarget{ref-lacal2018estimating}{}%
---------. 2019. ``Estimating and Testing Nonlinear Local Dependence Between Two Time Series.'' \emph{Journal of Business \& Economic Statistics} 37 (4): 648--60.
